---
title: "Normalization and Integration"
date: "2024-07-23"
output: 
  html_notebook:
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "", fig.align = "center", fig.height = 5, fig.width = 7, warning = F)
knitr::opts_knit$set(root.dir = "/home/lodato/mmiotto/misc/single_cell_bootcamp_unimi/") 

library(Seurat)
library(SeuratData)
library(ggplot2)
```

## Load data
```{r}
obj <- LoadData("pbmcsca")
obj
```

## Basic QC
We now filter cells with more than 1000 features.
```{r}
obj <- subset(obj, nFeature_RNA > 1000)
obj
```

The object contains data from nine different batches (stored in the Method column in the object metadata), representing seven different technologies.
We can see them:
```{r}
table(obj$Method)
```

## Normalization
The first step after the QC is normalization, to make gene expression comparable between cells. 

As we have multiple batches, we have first to split the object RNA layer into them, and then perform the normalization.

```{r}
obj[["RNA"]] <- split(obj[["RNA"]], f = obj$Method)
obj
```

We can now perform the normalization for each batch with just one line of code:
```{r}
obj <- NormalizeData(obj)
obj
```
This normalization set a scale factor of 10,000 for each cell (so the sum of the counts of a cell is set to 10,000), and adjust the expression value of each gene accordingly. It then calculate the log1p of those values (natural logarithm of the value+1 to account for zeros).

We can confirm that normalization has made the cells expression comparable:
```{r}
apply(obj[["RNA"]]$counts.10x_Chromium_v2[, 1:10], MARGIN = 2, FUN = function(x) {sum(x)})
```


```{r}
apply(obj[["RNA"]]$data.10x_Chromium_v2[, 1:10], MARGIN = 2, FUN = function(x) {sum(exp(x) - 1)})
```

## Dimensionality reduction
In order to run the PCA, we have to perform two operations on the data:

* Extrapolate variable features
* Scale data for those features: when performing PCA, input data should be scaled so that each feature has the same "weight"

```{r}
obj <- FindVariableFeatures(obj)
head(VariableFeatures(obj), 10)
```

```{r}
obj <- ScaleData(obj)
obj
```

### PCA
```{r}
obj <- RunPCA(obj)
obj
```

```{r}
DimPlot(obj, group.by = "Method")
```

```{r}
ElbowPlot(obj, ndims = 50) +
  scale_x_continuous(breaks = seq(0, 50, 5))
```


### UMAP
To better highlight differences in the sample, we can visualize the data with the UMAP.
```{r}
obj <- RunUMAP(obj, dims = 1:16, reduction = "pca", reduction.name = "umap.unintegrated")
obj
```

We can now visualize it:
```{r}
DimPlot(obj, reduction = "umap.unintegrated", group.by = c("Method"))
```

We can see that they clusterize by Method. This data should be integrated, as in scRNA seq experiment the clusters should be celltypes.

## Integration
To perform the integration of the data, we will use the `IntegrateLayers` funtion whithin Seurat. There are different integration methods: CCA, RPCA, Harmony, FastMNN and scVI. We will show CCA and Harmony.

### CCA
```{r}
obj <- IntegrateLayers(
  object = obj, method = CCAIntegration,
  orig.reduction = "pca", new.reduction = "integrated.cca",
  verbose = FALSE
)
```


```{r}
obj <- RunUMAP(obj, dims = 1:16, reduction = "integrated.cca", reduction.name = "umap.cca")
obj
```

```{r}
DimPlot(obj, reduction = "umap.cca", group.by = c("Method"))
```

### Harmony
```{r}
obj <- IntegrateLayers(
  object = obj, method = HarmonyIntegration,
  orig.reduction = "pca", new.reduction = "integrated.harmony",
  verbose = FALSE
)
```


```{r}
obj <- RunUMAP(obj, dims = 1:16, reduction = "integrated.harmony", reduction.name = "umap.harmony")
obj
```

```{r}
DimPlot(obj, reduction = "umap.harmony", group.by = c("Method"))
```

## Extra: SCTNormalization
There is also another normalization method in Seurat, which is an upgraded version of NormalizeData: `SCTransform`. It will replace NormalizeData, FindVariableFeatures and ScaleData functions.
It will introduce a new assay: SCT.

```{r}
obj <- SCTransform(obj, vst.flavor = "v2")
obj
```

```{r}
devtools::session_info()
```















